---
title: "RMS Titanic"
author: "Elias Tuthill"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>%
  glimpse()
```

**Observations**:

- Variables: Class, Sex, Age, Survived, n

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
  - According to Wikipedia, there were 2,208 aboard the Titanic. According to the dataset, there are a total number of 2,201 people.
- Are there any differences?
  - Yes, there are 7 fewer people in the dataset than there are according to the Titanic Wikipedia article.
- If yes, what might account for those differences?
  - It's likely that the records of the time were not perfect so it was hard to know which list of people is accurate. In fact, the Sinking of the Titanic Wikipedia article has a table that is different from the one on the Titanic page, and has 2,201 people on it. Between these tables, there's a lot more differences than just having 7 extra people. This reinforces the idea that no dataset of the Titanic passengers/crew is fully reliable.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(
    aes(
      x = Sex, 
      fill = Class,
      weight = n)
    ) +
  geom_bar()

```

**Observations**:

- It's hard to make any conclusions from this without comparing it to the total numbers of starting people in the various catagories or looking at survival rates. For example, there are slightly more male survivors than there are female, but the thing you always hear about the titanic is how the women and children were prioritized over men for lifeboats. However, it is likely given the time period that there were significantly more men on board the Titanic to start with, especially among the crew. This does come across in the survivor numbers among the crew, with there being significantly more men than women. I would guess that first class passengers were also prioritized for lifeboats but while the first class women have the largest number of survivors, the first class men don't. Interestingly, there's also a comparatively small number of second class men.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(
    aes(
      x = Sex, 
      fill = Class,
      weight = Prop)
  ) +
  geom_bar()
```

**Observations**:

- Write your observations here.
  - The plot shows that overall there are more female survivors than male. It also seems that first class dominates second which dominates third. Female crew looks roughly equal to third class whereas male is slightly less crew.
- Is there anything *fishy* going on in your plot?
  - Something fishy is going on with the Prop count. The count values don't make sense. It shouldn't be possible for there to be anything higher than 1 for any of the classes but it's very clear, especially looking at first class, that there is something higher than 1. Looking at the code, I have a suspicion that it has to do with the fact that age was factored into the grouping. Looking at the glimpse of the dataset, there's two categories of age: "Child" and "Adult". So I'm guessing that what is happening is that, for example with the female first class passengers, it's taking the child female first class passenger survivor proportion and adding it to the adult female first class passenger survivor proportion.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}

df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(
    aes(
      x = Sex, 
      fill = Class,
      weight = Prop)
  ) +
  geom_bar() +
  facet_grid(Age ~ .)
```

**Observations**:

- First thing I noticed was that very few adult men survived in comparison to all other categories, which tracks with the women and children first approach to getting people off the Titanic. Next is the helpful validation that there are no children in the crew. For the time it wouldn't have been impossible but I imagine it would've been fairly unlikely, at least for them to be labelled as children. Also, this continues to back up the class divide in survivorship. Though notably the first and second class are pretty close in all categories except adult men, which does leave me wondering why so few second class adult men survived. Perhaps they were the largest population?
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - This confirms the reason I suspected for the fishiness in q4. The prop was calculated for both age and sex but only plotted based on sex which meant the two proportions were added together. Looking back at q4, another interesting observation I've made about the fishy plot is that it seems like the proportion of crew to passenger survival is extremely low. This wasn't really something I would've questioned because it makes sense for something to happen like the crew to be involved in organizing passengers getting off, leading to fewer crew surviving. However, this plot shows that that isn't actually the case. Proportionally, significantly more adult female crew survived than adult female third class passengers. The reason the proportion wasn't inflated previously was because of the lack of crew children. That was the only proportion being accurately calculated.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
