---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!

gapminder %>%
  glimpse()
```

**Observations**:

- country, continent, year, lifeExxp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>%
    pull(year) %>%
    max()
year_min <- 
  gapminder %>%
    pull(year) %>%
    min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(continent, gdpPercap)) +
    geom_boxplot() +
    scale_y_log10()
    
```

**Observations**:

- African countries tend to have the lowest gdp per capita. Oceania's median is the highest. Asia has the largest spread. Asia has a major outlier with a significantly higher gdp per capita and the America's two less extreme outliers in the same direction.

**Difficulties & Approaches**:

- When I first ran the plot, it was very hard to read because Asia's massive outlier pushed the upper bound so high that everything else got majorly compressed. But adding a log scale made the plot readable.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2

Asia_outlier <-
  gapminder %>%
    filter(
      year == year_min,
      continent == "Asia", 
      gdpPercap >= 1e+05
      )
    
Americas_outliers <-
  gapminder %>%
    filter(
      year == year_min,
      continent == "Americas", 
      gdpPercap >= 1e+04
      )

Americas_outliers
Asia_outlier
```

**Observations**:

- Identify the outlier countries from q2
  - Canada, United States, Kuwait

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent


gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  {
    ggplot(., aes(x = continent, y = gdpPercap)) +
      geom_boxplot() +
      geom_point(
        data = . %>% filter(country %in% 
          c("Canada", "United States", "Kuwait", "Haiti")),
        aes(color = country)
      ) +
      scale_y_log10() +
      facet_wrap(~year)
  }
```

**Observations**:

- There is an additional outlier in 2007 that wasn't an outlier in 1952. I identified this as Haiti. It looks like a lot of countries have increased gdpPercap from 1972 to 2007. However it also looks like the overall spread has increased as well.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph

gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(size = pop, color = continent)) +
  geom_smooth(method = "lm") +
  scale_x_log10()
```

- I notice that Africa trends toward lower lifeExp and gdpPercap and Europe, Oceania, and Americas tend to be higher lifeExp and gdpPercap. Asia seems to be a decent spread. I also noticed there is a noticeable positive correlation and added a linear fit. I don't see much overall correlation between population and the linear fit. However I do notice that Asia's higher population countries tend to have lower gdpPercap and the Americas tend to have higher gdpPercap and lifeExp at higher populations. I also notice that when it comes to the linear fit, Asia tends to contribute a lot of outliers.

```{r q5-task2}
## TASK: Your second graph

# gap between biggest and smallest gdppercap
gapminder %>%
  group_by(year) %>%
  summarize(
    ratio_gdp = max(gdpPercap) / min(gdpPercap)
  ) %>%
  ggplot(aes(year, ratio_gdp)) +
  geom_line() +
  geom_point() +
  labs(title = "Gap Between Highest and Lowest GDP per Capita")  +
  theme(plot.title = element_text(hjust = 0.5))


#gap between biggest and smallest lifeexp
gapminder %>%
  group_by(year) %>%
  summarize(
    ratio_lifeexp = max(lifeExp) / min(lifeExp)
  ) %>%
  ggplot(aes(year, ratio_lifeexp)) +
  geom_line() +
  geom_point() +
  labs(title = "Gap Between Highest and Lowest Life Expectancy") +
  theme(plot.title = element_text(hjust = 0.5))

check <-
  gapminder %>%
  filter(year == 1977) %>%
  arrange(desc(gdpPercap))
#check

# rwanda lifeexp over time
gapminder %>%
  filter(country == "Rwanda") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  labs(title = "Rwanda Life Expectancy Over Time") +
  theme(plot.title = element_text(hjust = 0.5))
# rwandian genocide

gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot(aes(year, gdpPercap)) +
  geom_line() +
  labs(title = "Kuwait GDP per Capita Over Time") +
  theme(plot.title = element_text(hjust = 0.5))

gapminder %>%
  group_by(year) %>%
  summarize(
    min_gdp = min(gdpPercap)
  ) %>%
  ggplot(aes(year, min_gdp)) +
  geom_line() +
  geom_point() +
  labs(title = "Lowest GDP per Capita Over Time") +
  theme(plot.title = element_text(hjust = 0.5))

# "The large oil revenues of the 1970s left many private individuals with substantial funds at their disposal. These funds prompted a speculation boom in the official stock market in the mid-1970s that culminated in a crash in 1977."
# Wikipedia Souk Al-Manakh stock market crash


gapminder %>%
  filter(country == "Cambodia") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  labs(title = "Cambodia Life Expectancy Over Time") +
  theme(plot.title = element_text(hjust = 0.5))
# cambodian genocide
```

- I started out by looking at the gaps between highest and lowest countries in GDP per capita and life expentency. Overall, there appears to be an downward trend in both ratios. With these, I noticed 3 significant spikes in the difference. In the GDP per capita plot, I noticed one spike at 1972 followed by a sharp drop in 1977. In the life expectancy plot, I see two spikes in 1977 and 1992. I pulled up the data for each of these years and sorted by the relevant variable. From that, I was able to pull out outliers. I found low outliers for both of the life expectancy ratio drops and the gdp ratio drop, and a high outlier for the spike in gdp ratio.
- For life expectancy, I found that the spike in 1977 was caused by Cambodia. Cambodia has an overall upward trend in life expectancy, but there's a massive drop in 1977. From research, it appears that Cambodia had a genocide during 1977, which tracks with this finding. For the 1992 spike, I pulled out Rwanda as an outlier. It too has an overall upward trend in life expectancy, but it has an even larger drop than Cambodia, which tracks with the larger 1992 spike on the life expectancy ratio drop. From research, this aligns pretty well with the Rwandia genocide in 1994.
- For GDP per capita, I looked at the data and found that both the rise and fall aligned with Kuwait. In fact, the entire graph looks pretty well aligned. Based on this, I decided to look at the lowest GDP per capita every year, which shows very little variation in relation to Kuwait's variation. This backs up my intial thought that the GDP ratio is mostly driven by Kuwait. As for historical context that I identified, I found that it aligned with the Souk Al-Manakh stock market crash. Kuwait is a wealthy country because of oil exports, and there was a significant increase in wealth and a stock market boom in the early seventies, which was followed by a crash.

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  group_by(year, continent) %>%
  summarize(pop = sum(pop)) %>%
  ggplot(aes(year, pop, fill = continent)) +
  geom_area()
```

- I notice that the overall population of all included countries increased over time. I also notice that Asia makes up a considerable amount of the population. Oceania takes up less than I expected, however upon further research, I found that Indonesia and Malaysia are not included in Oceania but rather in Asia so this made this make sense. Europe seems to be the steadiest population-wise over time, with the Americas being a close second.
